How is experience transformed into memory?

### 2018-08-29 paper/main.tex
minor typo fixes to intro

How might one go about formally characterizing the ``gist'' of an experience?  Any given moment of an experience might derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g.\ ][]{LernEtal11}.  Therefore perhaps the timecourse of how an event unfolds might be critical to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g.\ the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences typically change gradually, they also occasionally change suddenly~\citep[e.g., walking through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into compressed memories.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.

Here we sought to examine how temporal dynamics of a ``naturalistic'' experience was reflected in participants' later memories of that experience.  We analyzed an open dataset which comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recalled an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode (and of participants' verbal recalls).  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and we use Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize the evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{topic trajectories} that describe how the experiences evolve over time.  In other words, the episode's topic trajectory is a formalization of its gist.  Under this framework, successful remembering entails verbally ``traversing'' the topic trajectory of the original episode, thereby reproducing the gist of the original episode.  In addition, comparing the shapes of the topic trajectories of the original episode and of participants' retelling of the episode reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We also identified a network of brain structures whose responses (as participants watched the episode) reflected the gist of the episode, and a second network whose responses reflected how participants would later recount the episode.

>>>>

How might one go about formally characterizing the ``gist'' of an experience?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore the timecourse of how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g.\ the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.  Although our experiences typically change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into compressed memories.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were reflected in participants' later memories of that experience.  We analyzed an open dataset which comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recalled an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode (and of participants' verbal recalls).  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and we use Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize the evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{topic trajectories} that describe how the experiences evolve over time.  In other words, the episode's topic trajectory is a formalization of its gist.  Under this framework, successful remembering entails verbally ``traversing'' the topic trajectory of the original episode, thereby reproducing the gist of the original episode.  In addition, comparing the shapes of the topic trajectories of the original episode and of participants' retellings of the episode reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We also identified a network of brain structures whose responses (as participants watched the episode) reflected the gist of the episode, and a second network whose responses reflected how participants would later recount the episode.
